The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-10-11 talk/vmil2012/presentation/talk.tex
a rough structure of the talk
fill the slides with a first rough draft

\date{2012 VMIL, XXX}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

@@

% XXX todos:
% note that two fields is a simplification
% have a diagram for the example trace
% lifting can never produce new operations
% have some sort of overview or position indicator for the many graphical slides, too easy to get lost
% more details about benchmarks, diagram?

% Structuring a talk is a difficult task and the following structure
% may not be suitable. Here are some rules that apply for this
% solution:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

% - A conference audience is likely to know very little of what you
%   are going to talk about. So *simplify*!
% - In a 20min talk, getting the main ideas across is hard
%   enough. Leave out details, even if it means being less precise than
%   you think necessary.
% - If you omit details that are vital to the proof/implementation,
%   just say so once. Everybody will be happy with that.

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \begin{itemize}
          \item for interesting, commonly executed code paths
          \item produces a linear list of operations (trace)
      \end{itemize}
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

\end{document}

>>>>

\date{2012 VMIL, 21st of October, 2012}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

@@

\section{Introduction}

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \begin{itemize}
          \item for interesting, commonly executed code paths
          \item produces a linear list of operations (trace)
          \item automatically does (potentially deep) inlining
      \end{itemize}
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Guards}
  \begin{itemize}
      \item Points of control flow divergence are marked with guards
      \item Operations that check whether conditions are still true
      \item When a guard fails, execution of the trace stops and continues in the interpreter
      \pause
      \begin{block}{Guard Characteristics}
          \begin{itemize}
              \item lots of them, up to 20\% guards
              \item most never fail
              \item costly to implement
          \end{itemize}
      \end{block}
  \end{itemize}
\end{frame}

% this talk wants to go over a lot of details that are usually glossed over as
% "easy" when tracing JITs are introduced.

\begin{frame}
  \frametitle{Bridges}
  \begin{itemize}
      \item When a trace fails often, it becomes worth to attach a new trace to it
          \item This is called a bridge
          \item The bridge is attached by patching the guard machine code
          \item when this guard fails in the future, the new trace is executed instead
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{RPython and PyPy}
  \begin{itemize}
      \item Context: RPython
      \item a generic tracing JIT, applicable to many languages
      \item main use: PyPy, an efficient Python interpreter
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Running Example}
\end{frame}

\section{High-Level}

\begin{frame}
  \frametitle{Symbolic Frame Capturing}
  \begin{itemize}
      \item Guard can fail deep inside inlined function
      \item when going back to the interpreter, call stack needs to be re-created
      \item done with the help of symbolic frame stacks
      \item these show how trace trace variables fill the to-be-built stack frames
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Symbolic Frame Compression}
  \begin{itemize}
      \item There are \emph{a lot of} guards
      \item Naively storing symbolic frames would be costly in terms of memory
      \item need to store them compactly
      \item observation: from one guard to the next, the non-top stack frames don't change
      \item share these between subsequent guards
      \pause
      \item also need a byte-saving binary representation, but that's just boring work
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Interaction with Optimization}
  \begin{itemize}
      \item Some optimizations make it necessary to store extra information in symbolic frames
      \item examples:
          \begin{itemize}
              \item delayed heap stores (need to do stores before resuming interpreter)
              \item allocation remvoal (need to allocate objects before resuming)
          \end{itemize}
      \item can be compressed using similar techniques
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Emitting Guards}
\end{frame}

\begin{frame}
  \frametitle{Patching Guards for Bridges}
\end{frame}

\section{Evaluation}

%as in paper
%fancy graphs
%something about execution speed
% <demo, if there is time>
\end{document}
